Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-11-16 mnras.tex
Converted to British spelling.

\documentclass[useAMS,usenatbib,a4paper]{mn2e}

@@

  We describe a photon-conserving radiative transfer algorithm, using
  a spatially-adaptive ray tracing scheme, and its parallel
  implementation into the adaptive mesh refinement (AMR) cosmological
  hydrodynamics code, Enzo.  By coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive time-stepping scheme and
  a fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  The test suite also
  includes an expanding \hii region in a magnetized medium, utilizing
  the newly implemented magnetohydrodynamics module in \enzo.  This
  method linearly scales with the number of point sources and number
  of grid cells.  Our implementation is scalable to 512 processors on
  distributed memory machines and can include radiation pressure and
  secondary ionizations from X-ray radiation.  It is included in the
  newest public release of Enzo.

@@

An important case was considered by \citet{Stroemgren39} for an
ultraviolet (UV) radiation source photoionizing a static uniform
neutral medium.  When recombinations balance photo-ionizations, the
radius of a so-called \hii region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_{\rm
        H}^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics of
the ISM.  A decade later, the first radiation hydrodynamical numerical
models of \hii regions in spherical symmetry and plane-parallel
ionization fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move supersonically and subsonically with respect to the
ionized gas, respectively.  The same terminology conversely applies to
D-type fronts.  ``Critical'' fronts are defined as moving exactly at
the sound speed.  These works established the evolutionary track of an
expanding \hii illuminated by a massive star in a uniform medium:
%
\begin{enumerate}
\item Weak R-type:  When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type:  As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type:  The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase:  After the shock forms, the \hii region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champagne flows.  Here the source is embedded
in an overdense region, and the \hii region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods: The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closure relation \citep{Gonzalez07, Aubert08}.
  Moment methods have the advantage of being fast and independent of
  the number of radiation sources.  However, they are diffusive and
  result in incorrect shadows in some situations.
\item Ray tracing:  Radiation can be propagated along rays that extend
  through the computational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparison paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

In this paper, we present our implementation, Enzo+Moray, of adaptive
ray tracing \citep{Abel02_RT} in the cosmological hydrodynamics
adaptive mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997,
  OShea2004}.  The radiation field is coupled to the hydrodynamics
solver at small timescales, enabling it to study radiation
hydrodynamical problems.  We have used this code to investigate the
growth of an \hii region from a 100\Ms~Population III (Pop III) star
\citep{Abel07}, the early stages of reionization from Pop III stars
\citep{Wise08_Reion}, radiative feedback on the formation of high
redshift dwarf galaxies \citep{Wise08_Gal, Wise10_Gal}, ultraviolet
radiation escape fractions from dwarf galaxies before reionization
\citep{Wise09}, negative radiative feedback from accreting Pop III
seed black holes \citep{Alvarez09}, and radiative feedback in
accreting supermassive black holes \citep[][in prep.]{Kim11}.  We have
included Enzo+Moray in the latest public release of
\enzo\footnote{http://enzo.googlecode.com}, and it is also coupled
with the newly added MHD solver in \enzo~\citep{Wang09}.

We have structured this paper as follows.  In Section 2, we describe
the mathematical connections between adaptive ray tracing and the
radiative transfer equation.  Furthermore, we detail how physics other
than photo-ionization and photo-heated are included.  We then derive a
geometric correction factor to any ray tracing method to improve
accuracy.  We end the section by describing a new computational
technique to approximate an optically-thin radiation field with ray
tracing and multiple sources.  In Section 3, we cover the details of
our radiation hydrodynamics implementation in \enzo, specifically (1)
the ray tracing algorithms, (2) coupling with the hydrodynamics
solver, (3) several methods to calculate the radiative transfer
timestep, and (4) our parallelization strategy.  We present our
results from the RT06 radiative transfer tests in Section 4.
Afterwards in Section 5, we show the results from the RT09 radiation
hydrodynamics tests.  We expand on these tests to include more
dynamical and complex setups to demonstrate the flexibility and high
fidelity of Enzo+Moray.  Section 6 gives the results from spatial,
angular, frequency, and temporal resolution tests.  In Section 7, we
illustrate the improvements from the geometric correction factor and
our optically-thin approximation.  We also show the effects of X-ray
radiation and radiation pressure in this section.  Finally in Section
8, we demonstrate the parallel scalability of Enzo+Moray.  Last
Section 9 summarizes our method and results.

@@

Solving this equation is difficult because of its high dimensionality;
however, we can make some appropriate approximations to reduce its
complexity in order to include radiation transport in numerical
calculations.  Typically timesteps in dynamic calculations are small
enough so that $\Delta a/a \ll 1$, therefore $\bar{a} = 1$ in any
given timestep, reducing the second term to $\hat{n} \partial
I_\nu/\partial \mathbf{x}$.  To determine the importance of the third
term, we evaluate the ratio of the third term to the second term.
This is $HL/c$, where $L$ is the simulation box length.  If this ratio
is $\ll 1$, we can ignore the third term.  For example at $z=5$, this
ratio is 0.1 when $L = c/H(z=5)$ = 53 proper Mpc.  In large boxes
where the light crossing time is comparable to the Hubble time, then
it becomes important to consider cosmological redshifting and dilution
of the radiation.  Thus equation (\ref{eqn:rteqn}) reduces to the
non-cosmological form in this local approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.  Next we describe this
discretization and its contribution to the radiation field.

@@

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (ph s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.

@@

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \; dr$,
and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centered density in our calculations.  Using trilinearly
interpolated densities \citep[see][]{Mellema06} did not produce
improved results.  Equation (\ref{eqn:rtray}) has a simple exponential
analytic solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionization state and
energy of the cells.  Considering a system with only hydrogen
photo-ionizations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\label{eqn:dndt}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \;
\mathrm{s}^{-1}$ is the recombination coefficient at 10$^4$ K in the
Case B on-the-spot approximation in which all recombinations are
locally reabsorbed, \citep{Spitzer78}, and $C_{\rm H}$ is the
collisional ionization rate.  However, for more accurate solutions in
calculations that consider several chemical species, the
photo-ionization rates are terms in the relevant chemical networks
\citep[e.g.][]{Abel97}.

@@

\begin{figure}
  \plotone{fig/geo_corr.eps}
  \caption{\label{fig:covering} (a) A two-dimensional illustration of
    the overlap between the beam associated with a ray $\gamma$ and a
    computational cell.  The ray has a segment length of $dr$ passing
    through the cell.  The covering area is denoted by dark grey,
    where the full area $(dr \times L_{\rm pix})$ is colored by the
    dark and light grey.  The photo-ionization and photo-heating rates
    should be corrected by the ratio of these areas, i.e. the overlap
    fraction $f_c$.  (b) Annotation of quantities used in this
    geometric correction.}
\end{figure}

For a ray tracing method to accurately, i.e. without non-spherical
artifacts, compute the radiation field, the computational grid must be
well-sampled by the rays.  The main source of potential artifacts is
the geometrical difference between the cell and the HEALPix pixel
relevant in the angular integration of the intensity over the cell.
In this section, we devise a correction scheme to account for these
differences.  Consider the solid angle $\Omega_{\rm ray}$ and photon
flux $P$ associated with a single ray, and assume the flux is constant
across $\Omega_{\rm ray}$.  There exists a discrepancy between the
geometry cell face and HEALPix pixel when the pixel does not cover the
entire cell face, which is illustrated in Fig. \ref{fig:covering}.
This mismatch causes non-spherical artifacts and is most apparent in
the optically thin case, where the area of the pixel is dominant over
$(1 - e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values ($>10$)
but we have formulated a simple geometric correction to the
calculation of the radiation field.  This correction is not unique to
the HEALPix formalism but can be applied to any type of pixelization.

>>>>

\documentclass[useAMS,usenatbib]{mn2e}

@@

  We describe a photon-conserving radiative transfer algorithm, using
  a spatially-adaptive ray tracing scheme, and its parallel
  implementation into the adaptive mesh refinement (AMR) cosmological
  hydrodynamics code, Enzo.  By coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilised to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive time-stepping scheme and
  a fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionisation
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  The test suite also
  includes an expanding \hii region in a magnetised medium, utilising
  the newly implemented magnetohydrodynamics module in \enzo.  This
  method linearly scales with the number of point sources and number
  of grid cells.  Our implementation is scalable to 512 processors on
  distributed memory machines and can include radiation pressure and
  secondary ionisations from X-ray radiation.  It is included in the
  newest public release of Enzo.

@@

An important case was considered by \citet{Stroemgren39} for an
ultraviolet (UV) radiation source photo-ionising a static uniform
neutral medium.  When recombinations balance photo-ionisations, the
radius of a so-called \hii region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_{\rm
        H}^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionising photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionised medium to be approximately the mean free path of
the ionising radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionising radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionisation front,
then this over-pressurised region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics of
the ISM.  A decade later, the first radiation hydrodynamical numerical
models of \hii regions in spherical symmetry and plane-parallel
ionisation fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionisation front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionisation fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionised gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move supersonically and subsonically with respect to the
ionised gas, respectively.  The same terminology conversely applies to
D-type fronts.  ``Critical'' fronts are defined as moving exactly at
the sound speed.  These works established the evolutionary track of an
expanding \hii illuminated by a massive star in a uniform medium:
%
\begin{enumerate}
\item Weak R-type:  When the star (gradually) starts to shine, the
  ionisation front will move supersonically through the ambient
  medium.  The gas is heated and ionised, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type:  As the ionisation front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type:  The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionised
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionisation front and form a shock.  The ionisation front
  detaches from the shock, putting the shock ahead of the ionisation
  front.
\item Expansion phase:  After the shock forms, the \hii region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionising the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champagne flows.  Here the source is embedded
in an overdense region, and the \hii region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionisation front passes this boundary, the dense, ionised gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods: The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closure relation \citep{Gonzalez07, Aubert08}.
  Moment methods have the advantage of being fast and independent of
  the number of radiation sources.  However, they are diffusive and
  result in incorrect shadows in some situations.
\item Ray tracing:  Radiation can be propagated along rays that extend
  through the computational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionisation by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparison paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

In this paper, we present our implementation, Enzo+Moray, of adaptive
ray tracing \citep{Abel02_RT} in the cosmological hydrodynamics
adaptive mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997,
  OShea2004}.  The radiation field is coupled to the hydrodynamics
solver at small time-scales, enabling it to study radiation
hydrodynamical problems.  We have used this code to investigate the
growth of an \hii region from a 100\Ms~Population III (Pop III) star
\citep{Abel07}, the early stages of reionisation from Pop III stars
\citep{Wise08_Reion}, radiative feedback on the formation of high
redshift dwarf galaxies \citep{Wise08_Gal, Wise10_Gal}, ultraviolet
radiation escape fractions from dwarf galaxies before reionisation
\citep{Wise09}, negative radiative feedback from accreting Pop III
seed black holes \citep{Alvarez09}, and radiative feedback in
accreting supermassive black holes \citep[][in prep.]{Kim11}.  We have
included Enzo+Moray in the latest public release of
\enzo\footnote{http://enzo.googlecode.com}, and it is also coupled
with the newly added MHD solver in \enzo~\citep{Wang09}.

We have structured this paper as follows.  In Section 2, we describe
the mathematical connections between adaptive ray tracing and the
radiative transfer equation.  Furthermore, we detail how physics other
than photo-ionisation and photo-heated are included.  We then derive a
geometric correction factor to any ray tracing method to improve
accuracy.  We end the section by describing a new computational
technique to approximate an optically-thin radiation field with ray
tracing and multiple sources.  In Section 3, we cover the details of
our radiation hydrodynamics implementation in \enzo, specifically (1)
the ray tracing algorithms, (2) coupling with the hydrodynamics
solver, (3) several methods to calculate the radiative transfer
timestep, and (4) our parallelisation strategy.  We present our
results from the RT06 radiative transfer tests in Section 4.
Afterwards in Section 5, we show the results from the RT09 radiation
hydrodynamics tests.  We expand on these tests to include more
dynamical and complex setups to demonstrate the flexibility and high
fidelity of Enzo+Moray.  Section 6 gives the results from spatial,
angular, frequency, and temporal resolution tests.  In Section 7, we
illustrate the improvements from the geometric correction factor and
our optically-thin approximation.  We also show the effects of X-ray
radiation and radiation pressure in this section.  Finally in Section
8, we demonstrate the parallel scalability of Enzo+Moray.  Last
Section 9 summarises our method and results.

@@

Solving this equation is difficult because of its high dimensionality;
however, we can make some appropriate approximations to reduce its
complexity in order to include radiation transport in numerical
calculations.  Typically timesteps in dynamic calculations are small
enough so that $\Delta a/a \ll 1$, therefore $\bar{a} = 1$ in any
given timestep, reducing the second term to $\hat{n} \partial
I_\nu/\partial \mathbf{x}$.  To determine the importance of the third
term, we evaluate the ratio of the third term to the second term.
This is $HL/c$, where $L$ is the simulation box length.  If this ratio
is $\ll 1$, we can ignore the third term.  For example at $z=5$, this
ratio is 0.1 when $L = c/H(z=5)$ = 53 proper Mpc.  In large boxes
where the light crossing time is comparable to the Hubble time, then
it becomes important to consider cosmological redshifting and dilution
of the radiation.  Thus equation (\ref{eqn:rteqn}) reduces to the
non-cosmological form in this local approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.  Next we describe this
discretisation and its contribution to the radiation field.

@@

We avoid this inefficiency by utilising adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centres of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialised at
each point source with the photon luminosity (ph s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionisation state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.

@@

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretisation of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \; dr$,
and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centred density in our calculations.  Using trilinearly
interpolated densities \citep[see][]{Mellema06} did not produce
improved results.  Equation (\ref{eqn:rtray}) has a simple exponential
analytic solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionisation rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionisation $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionisation energy of the absorbing material.
In each cell, the photo-ionisation and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionisation state and
energy of the cells.  Considering a system with only hydrogen
photo-ionisations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\label{eqn:dndt}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \;
\mathrm{s}^{-1}$ is the recombination coefficient at 10$^4$ K in the
Case B on-the-spot approximation in which all recombinations are
locally reabsorbed, \citep{Spitzer78}, and $C_{\rm H}$ is the
collisional ionisation rate.  However, for more accurate solutions in
calculations that consider several chemical species, the
photo-ionisation rates are terms in the relevant chemical networks
\citep[e.g.][]{Abel97}.

@@

\begin{figure}
  \plotone{fig/geo_corr.eps}
  \caption{\label{fig:covering} (a) A two-dimensional illustration of
    the overlap between the beam associated with a ray $\gamma$ and a
    computational cell.  The ray has a segment length of $dr$ passing
    through the cell.  The covering area is denoted by dark grey,
    where the full area $(dr \times L_{\rm pix})$ is colored by the
    dark and light grey.  The photo-ionisation and photo-heating rates
    should be corrected by the ratio of these areas, i.e. the overlap
    fraction $f_c$.  (b) Annotation of quantities used in this
    geometric correction.}
\end{figure}

For a ray tracing method to accurately, i.e. without non-spherical
artifacts, compute the radiation field, the computational grid must be
well-sampled by the rays.  The main source of potential artifacts is
the geometrical difference between the cell and the HEALPix pixel
relevant in the angular integration of the intensity over the cell.
In this section, we devise a correction scheme to account for these
differences.  Consider the solid angle $\Omega_{\rm ray}$ and photon
flux $P$ associated with a single ray, and assume the flux is constant
across $\Omega_{\rm ray}$.  There exists a discrepancy between the
geometry cell face and HEALPix pixel when the pixel does not cover the
entire cell face, which is illustrated in Fig. \ref{fig:covering}.
This mismatch causes non-spherical artifacts and is most apparent in
the optically thin case, where the area of the pixel is dominant over
$(1 - e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values ($>10$)
but we have formulated a simple geometric correction to the
calculation of the radiation field.  This correction is not unique to
the HEALPix formalism but can be applied to any type of pixelisation.
